\documentclass[12pt]{article}
\usepackage[margin=0.25in,letterpaper]{geometry}
\usepackage[ampersand]{easylist}

%% or use amsymb for open and closed items
%% \usepackage{amssymb}
%% \begin{itemize}
%% \item[$\square$] An open item.
%% \item[$\boxtimes$] A closed item.
%% \end{itemize}

\title{Midterm 1 Preparation Checklist for \\ \normalsize STAT 445/645 Intro to Statistical Computing}
\author{
        AG Schissler
%%                Department of Mathematics \& Statistics\\
%%         University of Nevada, Reno
}
\date{\today}



\begin{document}
\maketitle


%% Please use this checklist~\cite{Gawande2010} to maximize your test performance. I recommend adding and sharing your own ideas for this list.
Please use this checklist to maximize your test performance. I recommend adding and sharing your own ideas for this list.

\section*{Checklist}

\begin{easylist}[checklist]
& 1.~Lab review
&& Reviewed Lab 01 and made notes.
&& Reviewed Lab 02 and made notes.
&& Reviewed Lab 03 and made notes.
& 2.~DataCamp review
&& Reviewed DataCamp Intro to R and made notes.
&& Reviewed DataCamp Intermediate R and made notes.
&& Reviewed DataCamp Writing functions in R and made notes.
&& Reviewed DataCamp Writing efficient R code and made notes.
&& Reviewed DataCamp Intermediate R - Practice and made notes.
& 3.~Lecture review
&& Reviewed lesson1 intro and made notes.
&& Reviewed lesson2 data structures and made notes.
&& Reviewed lesson3 flow and made notes.
&& Reviewed lesson4 lab01 and made notes.
&& Reviewed lesson5 functions iteration and made notes.
&& Reviewed functions and made notes.
&& Reviewed lesson8 ttests text parallel and made notes.
&& Reviewed lesson10 midterm1 review and made notes.
& 4.~Support
&& Googled any questions that I have.
&& Asked classmates any questions that I have.
&& Asked the instructor any questions that I have.
& 5.~1-page notes preparation
&& Summarized my notes on 1 8.5inx11in paper (front and back)
&& Reviewed my paper so that I know where everything is.
& 6.~Prep for test
&& Ensure I get a good night's sleep before the test.
&& Bring pencils, blue or black pen, and scratch paper.
&& Bring a silent timer monitor time without distraction during the test.
& 7.~During the test
&& Review the entire test and ask the instructor if unclear on anything.
&& Set a timer to give yourself a 15 or 20 minute warning.
&& Work at a steady, relaxed pace.
&& Answer every question - leave nothing blank.
\end{easylist}

%% \bibliographystyle{abbrv}
%% 
%% \begin{thebibliography}{10}
%% 
%% \bibitem{Gawande2010}
%% Gawande, Atul (2010) {\em The Checklist Manifesto}, vol.~1. Penguin Books India
%% 
%% \end{thebibliography}
%% 

\end{document}
