---
title: "Intro to Stat Computing Proposal Talks"
author: "AG Schissler"
date: "10/02/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

<bring to class: textbook>

## Start-of-class work: ( < 1 min)

Please turn take your seats and engage/participate in the talks. Hold Q & A until the end.

## I. Project proposal talks (50 - 70 minutes)

```{r prez_order, results='asis'}
group_dat <- readxl::read_xlsx("~/OneDrive - University of Nevada, Reno/Teaching/STAT_445-645/FA19_student_work/grad_project_groups_milestones.xlsx", sheet = "grad_project_groups_milestones")
## treat as categorical variable
group_dat$GroupID <- factor(group_dat$GroupID)
## randomize prez order
set.seed(10022019)
prez_order <- sample(levels(group_dat$GroupID))
## print(as.numeric(prez_order))
## print the names in that order
student_by_id_list <- split(group_dat$Student, group_dat$GroupID)
student_vec <- unlist(lapply(student_by_id_list, paste, collapse = " and "))
## print in prez order
topic_by_id_list <- split(group_dat$Topic, group_dat$GroupID)
topic_vec <- unlist(lapply(topic_by_id_list, function(x) x[1]))
cat(paste( 1:length(student_vec), ". **", student_vec[prez_order], "**  \n\n*", topic_vec[prez_order], "*  \n\n", sep = ""))
```


## Closing (5 min)

- Which project are you most interested in?
- 445 students please inquiry with groups as soon as possible if you'd like to participate in a project. Note that it is ultimately the graduate team's decision and please do not be offended if for whatever reason it doesn't work.
