---
title: "Week 7 / Lab 04: Reading and importing data"
author: "AG Schissler"
date: "Monday September 23, 2019"
output: html_document
---

## I. Opening 

```{r, include=FALSE}
knitr::opts_chunk$set(cache=TRUE, autodep=TRUE, cache.comments=TRUE)
```

### I. Start-of-class work

1. Log in the your workstation.
2. Open the RStudio application on your machine.
3. Open [12_importing_data.Rmd](https://github.com/grizant/STAT445/blob/master/meeting_agendas/12_importing_data.Rmd) from website so that you can work along the lecture.
4. Open [lab_04_reading_cleaning_data.Rmd](https://github.com/grizant/STAT445/blob/master/labs/lab_04_reading_cleaning_data.Rmd) to have the lab available for review.

### This week's agenda

- Begin *Working with data* Unit II.

### Today learning outcomes

Students will be able to

- read in (import) data into R
- write data to a file from R
- apply in to real data in lab 04

## II. Reading/writing data

All along, we've already been reading in data from the outside, using:

- `readLines()`: reading in lines of text from a file or webpage; returns vector of strings
- `read.table()`: read in a data table from a file or webpage; returns data frame
- `read.csv()`: like the above, but assumes comma separated data; returns data frame

This week we'll focus on `read.table()`, `read.csv()` and their counterparts `write.table()`, `write.csv()`, respectively.

#### Reading in data from a previous R session

- `readRDS()`, `saveRDS()`: functions for reading/writing single R objects from/to a file
- `load()`, `save()`: functions for reading/writing any number of R objects from/to a file

Advantage: these can be a lot more memory efficient than what we'll cover this week.
Disadvantage: they're limited in scope in the sense that they can only communicate with R.

#### `read.table()`, `read.csv()`

Have a table full of data, just not in the R file format? Then `read.table()` is the function for you. It works as in:

- `read.table(file=file_name, sep=" ")`, to read data from a local file on your computer called `file_name`, assuming (say) space separated data
- `read.table(file=webpage_link, sep="\t")`, to read data from a webpage up at `webpage_link`, assuming (say) tab separated data

The function `read.csv()` is just a shortcut for using `read.table()` with `sep=","`. (But note: these two actually differ on some of the default inputs!)

#### Helpful input arguments

The following inputs apply to either `read.table()` or `read.csv()` (though these two functions actually have different default inputs in general---e.g., `header` defaults to TRUE in `read.csv()` but FALSE in `read.table()`)

- `header`: boolean, TRUE is the first line should be interpreted as column names
- `sep`: string, specifies what separates the entries; empty string "" is interpreted to mean any white space
- `quote`: string, specifies what set of characters signify the beginning and end of quotes; empty string "" disables quotes altogether

Other helpful inputs: `skip`, `row.names`, `col.names`. You can read about them in the help file for `read.table()` .

Another example of reading in data
===

```{r, error=TRUE}
# This data table is tab separated, and it comes with column names
sprint_df <- read.table(
  file="http://www.stat.cmu.edu/~ryantibs/statcomp/data/sprint.m.dat",
  sep="\t", header=TRUE)

# Oops! It turns out it has some apostrophe marks in the City column, 
# so we have to set quote=""
sprint_df <- read.table(
  file="http://www.stat.cmu.edu/~ryantibs/statcomp/data/sprint.m.dat",
  sep="\t", header=TRUE, quote="")
head(sprint_df)
sapply(sprint_df, class)
unique(grep("'", sprint_df$City, value=TRUE)) # This is the troublemaker
```

#### `write.table()`, `write.csv()`

To write a data frame (or matrix) to a text file, use `write.table()` or `write.csv()`. These are the counterparts to `read.table()` and `read.csv()`, and they work as in:

- `write.table(my_df, file="my_df.txt", sep=" ", quote=FALSE)`, to write `my_dat` to the text file "my_df_txt" (to be created in your working directory), with (say) space separation, and no quotes around the entries
- `write.csv(my_df, file="my_df_csv", quote=FALSE)`, to write `my_dat` to the text file "my_df.csv" (to be created in your working directory), with comma separation, and no quotes.

Note that `quote=FALSE`, signifying that no quotes should be put around the printed data entries, seems always preferable (default is `quote=TRUE`). Also, setting `row.names=FALSE` and `col.names=FALSE` will disable the printing of row and column names (defaults are `row.names=TRUE` and `col.names=TRUE`).


## III. Lab 04

1. Review the entire document.
2. Ask for clarification as needed.
3. Identify next actions.

## IV. Summary

- 3 new R functions
- 2 stats ideas
- 1 action item for you to do.
