---
title: "Week 8 / Exploratory Data Analysis, numeric summary, Lab 05"
author: "AG Schissler"
date: "Wednesday October 16, 2019"
output: html_document
---

## I. Opening 

```{r, include=FALSE}
knitr::opts_chunk$set(cache=TRUE, autodep=TRUE, cache.comments=TRUE)
```

### I. Start-of-class work

1. Log in the your workstation.
2. Open the RStudio application on your machine.
3. Open [14_eda_numeric_startLab05.Rmd](https://github.com/grizant/STAT445/blob/master/meeting_agendas/14_eda_numeric_startLab05.Rmd) from website so that you can work along the lecture.
4. Open [lab_05_numeric_explore_apply.Rmd](https://github.com/grizant/STAT445/blob/master/labs/lab_05_numeric_explore_apply.Rmd) to have the lab available.

### This week's agenda

- Continue *Working with data* Unit II.
- Focus on using R to perform numeric exploratory data analysis (EDA).
- More work with the `apply` family of functions.

### Today learning outcomes

Students will be able to

- more practice on reading in (import) data into R
- use `apply` to numerically summarize real data in lab 05

## II. An outlier generating example through a discrete mixture

Let's discuss how to simulate data with outliers. One way this could arise is that the data in one variable are actually from two distributions that are *mixed* together.  

In the code below we simulate $n=1000$ data points from a discrete mixture of Gaussian random variables. Formally, let  

$$
X \sim N(0,1)
$$

and  

$$
Y \sim N(5,1)
$$

Then let  

$$
Z = X, \pi = 1
Z = Y, \pi = 0
$$ 


be the random variable that results from the (two-component) discrete mixing of $X$ and $Y$ with mixing proportion $\pi \sim bernoulli(p = 0.01)$. By the values we've used above the data will appear to have outliers --- a few points from away from the majority of points.

```{r, fig.cap = 'Figure 1. Histogram of 1000 simulated points from a discrete mixture of Gaussians. This produces a few outliers with larger values than the majority of points.'}
## let's simulate and visualize data with outlying points
set.seed(10162109)
## mixing probability
my_pi <- 0.01
## suppose with have 1000 data points
n = 1000
## let's sample which points will be from X and Y
my_index <- rbinom(n = n, size = 1, prob = my_pi)
## count the "outliers"
paste("There are", sum(my_index), "outliers.")
## simulate the values by an "on-the-fly"
## function definition in sapply
## development parameter: tmp_index <- 0
my_sim_data <- sapply(X = my_index, FUN = function(tmp_index){
    ## ifelse works since 1 == TRUE and 0 == FALSE
    ifelse(test = tmp_index, yes = rnorm(1, 5, 1), no = rnorm(1,0,1))
})
ggplot2::qplot(x = my_sim_data, bins = 40) + ggplot2::theme_bw() + ggplot2::labs(x="Simulated values", y = "Counts") 
```

## III. Please continue Lab 04

- Questions for the whole group? Please don't be shy it could save us all effort.
- Please begin and I will answer questions in the order I receive them and considering who I have helped previously in class.

## IV. Summary

Why do we perform EDA before fitting any statistical model?
