%A simple LaTeX syllabus template by Luke Smith.
%I believe this is a variation of a template I got by Mike Hammond...?

%\documentclass[11pt,twocolumn]{article}
\documentclass[11pt,onecolumn]{article}
\usepackage{hyperref}
\usepackage[margin=0.75in]{geometry}
\usepackage{titlesec}
\usepackage{longtable}
\usepackage{gb4e}
%\usepackage[margin=.75in]{geometry}
\usepackage{textcomp}
\pagenumbering{gobble}
%% \usepackage[utf8]{inputenc}
%% \usepackage[english]{babel}
%% \usepackage[usenames, dvipsnames]{color}

%paragraph formatting
\setlength{\parindent}{0pt}
\setlength{\parskip}{3pt}

\usepackage{url}

\hypersetup{
    colorlinks=true,
    linkcolor=blue,
    filecolor=magenta,      
    urlcolor=blue,
}

\titlespacing\subsection{0in}{\parskip}{\parskip}
\titlespacing\section{0in}{\parskip}{\parskip}

%Just fill these in to fill in the basic syllabus information.
\newcommand{\coursename}{Introduction to Statistical Computing - STAT 445/645}
\newcommand{\semester}{Fall 2018}
\newcommand{\roomnumb}{DMSC 106}
\newcommand{\classtimes}{Mon,Wed 2:30pm - 3:45pm}
\newcommand{\myname}{A.~Grant Schissler}
%\newcommand{\myemail}{\href{mailto:aschissler@unr.edu}{aschissler@unr.edu}}
\newcommand{\myemail}{aschissler@unr.edu}
\newcommand{\office}{DMSC 224}
\newcommand{\officehours}{Tue 2:30pm-3:30pm, Wed 1:30pm-2:30pm, or by appointment}
% \newcommand{\university}{The University of Nevada, Reno}
\renewcommand{\familydefault}{\sfdefault}

\title{\textbf{\coursename}}
% \author{{\university}---{\semester}---{\roomnumb}---{\classtimes}}
\author{{\semester}---{\roomnumb}---{\classtimes}}
\date{}


\begin{document}
\maketitle

\vspace{-0.25in}
\noindent\makebox[\linewidth]{\rule{\textwidth}{1pt}}

\begin{center}
\begin{tabular}{llll}
\textbf{Instructor}:&\myname & \textbf{Contact}:&\href{mailto:\myemail}{\myemail}, 775-784-4661 (office)\\
\textbf{Office}:&\office & \textbf{Hours}:&\officehours\\
\end{tabular}
\end{center}

\section*{Catalog Description}
Introduction to statistical computing; data visualization and manipulation; document creation; graphics; simulation techniques; parallel computing; estimation; optimization; advanced statistical methods.

\section*{400-level Student Learning Outcomes}

\begin{description}
\item[UG1] Students will be able to implement statistical simulation, re-sampling techniques, and maximum likelihood estimation. 
\item[UG2] Students will be able to conduct a simulation-based power analysis. 
\item[UG3] Students will be able to write professional quality reports and computer code.
\item[GRAD1] Students will be able to use statistical computing methods to complete a research project and effectively communicate their findings.
\end{description}

\section*{Course outcomes}

Students will be able to $\ldots$

\begin{enumerate}
  \itemsep0em 
\item use R and RStudio
\item use basic git commands
\item produce HTML, PDF, or Word documents using R Markdown
\item install R packages for computing tasks
\item use vectors.
\item use lists.
\item use \textsc{data.frames}
\item apply data types (factor, numeric, etc.)
\item explore data structures using \textsc{str, head, tail, etc.}
\item use indexing and iteration to complete computing tasks
\item manipulate text in R (splitting, searching, cleaning)
\item use the \textsc{apply} family of functions to compute efficiently
\item demonstrate mastery of above content in Midterm 1
\item plot data using 1D, 2D, 3D strategies using \textsc{base,graphics}
\item apply the grammar of graphics through \textsc{ggplot2}
\item write R functions
\item simulate data using pseudo random-number generation
\item import/export data in various formats
\item demonstrate mastery of all above content in Midterm 2
\item apply the split-apply-combine data-analysis strategy
\item perform parallel computing tasks (parallelization)
\item fit basic statistical models and make predictions
\item integrate functions using Monte Carlo techniques
\item use re-sampling for statistical inference (bootstrap, jackknife)
\item use maximum likelihood estimation for statistical inference
\item use Markov Chain Monte Carlo (MCMC) to sample from probability distributions
\item demonstrate mastery of all topics in a Final Exam
\end{enumerate}


\end{document}