---
title: "Iteration and functions in R, lab_01 con'd"
author: "AG Schissler"
date: "9/12/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## I. Opening 

### Start-of-class work

1. Log in the your workstation.
2. Open the RStudio application on your machine.
3. Open your lab_01.Rmd.

### Today learning outcomes

Students will be able to

- use the `apply` family of R functions.
- compare efficiency of R code.
- compare the probability theory to simulated results
- use R lists, matrices, and vectors.
- use and write R functions.

## II. Recall Binomial random variable properties

Recall some properties for a binomial random variable:

$X \sim Bin(n,p)$  
$\mu = np$  
$\sigma^2 = np(1-p)$  
$P(k) = P(X = k) = {n \choose k} p^k(1-p)^{n-k}$  

**Hint** use `dbinom` to compute the pmf above.

## III. Sample variance versus population variance

The variance formula above is for a population, not the sample variance that serves an estimator to the population.

The `var` and `sd` functions in R and the unbiased sample estimators that use $n-1$ instead of $n$. 

$\hat{var(X)} = 1/(n-1) * \sum_{i=1}^{n} (x_i - \bar{x})^2$

To make the theoretical comparisons in the lab you may want to adjust by the following formula

```{r population_sd, eval=FALSE}
sd_pop <- sqrt( (n-1)/n ) * sd(x)
```

## IV. Clocking R code

You're asked to make several time comparisons. You may like to use the `system.time` function.

```{r clocking}
system.time(rnorm(n=1e6))
```

## V. 7c

There's a typo in Q7c you should use the function `standardize_slow` not just `standardize`.

## VI. Lab grading rubric

I'm developing a rubric for how your lab will be graded. I anticipate having that done tomorrow. Please review and let me know if you have any questions. I'll also post solutions in WebCampus Monday to give you feedback.

## VI. Closing


