---
title: "Control flow in R"
author: "AG Schissler"
date: "8/29/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## I. Opening:

### Start-of-class work

1. Log in the your workstation.
2. Open the RStudio application on your machine.

### Today learning outcomes

1. Students will be able to use R help functions.
2. Students will be able to control flow in R using conditionals and other tools.


## II. Intro to control flow in R

Control flow dictates the order/which operations are conducted.

### I do
Let's look at some R documentation on the topic.
These are functions you can use access R's documentation (instead of Googling).
```{r help}
help("Control")
```
Or you could use `?"Control"`.

### We do
There are other ways to search including fuzzy matching. Try running `help("control")`.
No match so let's search.

```{r help2, eval=F}
??"control"
## Or
help.search("control")
```
### You do
Look up documentaion on `t.test`. What package houses the code for that function?

## III. DataCamp: Intermediate R module 1

### You do
- Please work on the first module Conditional and Control Flow in DataCamp.
- Please make use of the _people_ in the classroom: ask **questions** of your substitute instructor and classmates.
- Take notes while learning by either copying DataCamp code into R scripts or hand notes. 

## IV. Closing

### Think-pair-share
1. Create a one-sentence summary of today's lesson.
2. Pair up with your neighbor and revise.
3. Share a few summaries with whole group.
