%A simple LaTeX syllabus template by Luke Smith.
%I believe this is a variation of a template I got by Mike Hammond...?

%\documentclass[11pt,twocolumn]{article}
\documentclass[11pt,onecolumn]{article}
\usepackage{hyperref}
\usepackage[margin=0.75in]{geometry}
\usepackage{titlesec}
\usepackage{longtable}
\usepackage{gb4e}
%\usepackage[margin=.75in]{geometry}
\usepackage{textcomp}
\pagenumbering{gobble}
%% \usepackage[utf8]{inputenc}
%% \usepackage[english]{babel}
%% \usepackage[usenames, dvipsnames]{color}

%paragraph formatting
\setlength{\parindent}{0pt}
\setlength{\parskip}{3pt}

\usepackage{url}

\hypersetup{
    colorlinks=true,
    linkcolor=blue,
    filecolor=magenta,      
    urlcolor=blue,
}

\titlespacing\subsection{0in}{\parskip}{\parskip}
\titlespacing\section{0in}{\parskip}{\parskip}

%Just fill these in to fill in the basic syllabus information.
\newcommand{\coursename}{Introduction to Statistical Computing - STAT 445/645}
\newcommand{\semester}{Fall 2018}
\newcommand{\roomnumb}{DMSC 106}
\newcommand{\classtimes}{Mon,Wed 2:30pm - 3:45pm}
\newcommand{\myname}{A.~Grant Schissler}
%\newcommand{\myemail}{\href{mailto:aschissler@unr.edu}{aschissler@unr.edu}}
\newcommand{\myemail}{aschissler@unr.edu}
\newcommand{\office}{DMSC 224}
\newcommand{\officehours}{Tue 2:30pm-3:30pm, Wed 1:30pm-2:30pm, or by appointment}
% \newcommand{\university}{The University of Nevada, Reno}
\renewcommand{\familydefault}{\sfdefault}

\title{\textbf{\coursename}}
% \author{{\university}---{\semester}---{\roomnumb}---{\classtimes}}
\author{{\semester}---{\roomnumb}---{\classtimes}}
\date{}


\begin{document}
\maketitle

\vspace{-0.25in}
\noindent\makebox[\linewidth]{\rule{\textwidth}{1pt}}

\begin{center}
\begin{tabular}{llll}
\textbf{Instructor}:&\myname & \textbf{Contact}:&\href{mailto:\myemail}{\myemail}, 775-784-4661 (office)\\
\textbf{Office}:&\office & \textbf{Hours}:&\officehours\\
\end{tabular}
\end{center}

Computational skills have become an invaluable asset in our increasingly quantitative world. This course introduces students to the foundational skills and concepts needed in modern statistical and mathematical computing, with an emphasis on high-level programming languages and statistical applications. 

There are three main parts to the course: First, students learn basic programming skills, graphics, and how to carry out basic statistical analyses (e.g., regression) all in \textsc{R}. Second, students learn the fundamentals of simulation techniques (Monte Carlo) for generating synthetic data. Third, students are introduced to various topics in computational statistics for estimation, integration, hypothesis testing, and density estimation for univariate and multivariate models.

Additionally, we'll briefly cover high-performance computing and document creation. This include parallel computing on desktops and/or on the Math server Okapi and the use \textsc{R} Markdown and \LaTeX. This course will primarily use \textsc{R}, but may introduce some \textsc{bash} shell (\textsc{Linux}) commands/scripting and Bayesian computation in \textsc{Stan}.

\section*{Catalog Description}
Introduction to statistical computing; data visualization and manipulation; document creation; graphics; simulation techniques; parallel computing; estimation; optimization; advanced statistical methods.

\section*{Course Pre-requisites}
STAT 352 or STAT 467/667 or with instructor approval.

\section*{400-level Student Learning Outcomes}
%% Student Learning Outcomes
%% Upon completion of this course, students will be able to:
%% 1. implement statistical simulation and resampling techniques, maximum likelihood estimation. 
%% 2. conduct a simulation-based power analysis. 
%% 3. write professional quality reports and computer code. 
\begin{description}
\item[UG1] Implement statistical simulation and resampling techniques, maximum likelihood estimation. 
\item[UG2] Conduct a simulation-based power analysis. 
\item[UG3] Write professional quality reports and computer code. 
\end{description}

\section*{600-level Student Learning Outcomes}
\begin{description}
\item[GRAD1] Complete a research project and effectively communicate their findings.
\end{description}

\clearpage
\section*{Textbooks}
None required. Supplementary texts:
\begin{itemize} 
\item Rizzo, M. \emph{Statistical Computing With R}. 2008, Chapman \& Hall/CRC.
%% $Textbook website:\\~\url{https://www.wiley.com/en-us/Introduction+to+Bayesian+Statistics%2C+3rd+Edition-p-9781118091562}
\item Grolemund, G., Wickham, H. \emph{R for Data Science}. 2016, O’Reilly.(Free at ~\url{http://r4ds.had.co.nz/})
\item The R Cookbook, by Paul Teetor
\item The R Graphics Cookbook, by Winston Chang
\item The Art of R Programming: A Tour of Statistical Software Design, by Norman Matloff
\item Data Manipulation with R, by Phil Spector
\item Software for Data Analysis: Programming with R, by John Chambers (advanced book)
\end{itemize}


\section*{Online resources}
\begin{itemize}
\item Course website:~\url{http:www.grantschissler.com/teaching/FA18/STAT445}, includes a working course schedule and approximate due dates of assignments and exams.
\item Course github repo:~\url{http:github.com/grizant/STAT445}
\item Students are responsible for checking Web Campus (\url{http:wcl.unr.edu}) and their email accounts, and are assumed to be aware of all information posted to these sources prior to each meeting.
\end{itemize}

\section*{Assignments}
Exercises will be assigned approximately weekly. You are encouraged to discuss assignments between each other and with instructor. However, the works must be completed and submitted individually.

\section*{Exam policy} You will be allowed at one 8.5x11in page of handwritten (on both sides) notes for each midterm and three such pages for the final exam. If you believe that your grade for exam or assignment is incorrect, contact instructor at the office hours with a rational justification. All such requests must be submitted to instructor within one week after a grade is announced; late requests will not be granted. The final decision about new grade is made by the instructor. Please understand that everyone can make a mistake, and that mistakes can go both ways: higher or lower than original grade.

\section*{Midterms}
There will be two midterms, the first on Wednesday, September 26, and the second on Wednesday, October 31.

\section*{Final exam}
A comprehensive final examination will be held on Wednesday, December 19, 12:10pm - 2:10pm, DMS 106.

\section*{400/600 Students}
As indicated above, the student learning outcomes differ at the 400 and 600 levels. Assignments and exams for students enrolled at the 400 level and 600 level will also differ. Specifically, graduate students will be expected to complete more exercises and exercises of greater difficulty on assignments. Exams will be of higher difficulty. Students enrolled at the 600 level will also be required to complete a \textbf{term project} due December 12.

\section*{Makeup, Late Policy}
Late assignments, exams, and projects will not be graded. Exceptions will be made when a student misses work due to documented (doctor's note) illness or other extraordinary situation, up to the discretion of the instructor. There will be no early or make-up exams. However, if you need to miss an exam due to participation in official university activities (including athletics and other sanctioned activities), you must make arrangements with the instructor at least two weeks prior to the exam in question. Since the late policy is rather strict, I will drop your lowest two grades in the ``Assignments'' category as a safety factor for emergencies.

\section*{Grading}
The final grades will be determined using the following percentages:

\begin{center}
\begin{tabular}{cc}
\begin{tabular}{l|c|r}	%For grade items (quizzes, homework, etc.)
Item&400-level& 600-level\\\hline\hline
  Assignments&40\% & 40\%\\
  Midterm Exams&40\%& 20\%\\
  Final Exam&20\%& 20\%\\
  Term project &-- &20\%\\
\end{tabular}
&
\begin{tabular}{ll}
A&90-100\\
B&80-89\\
C&70-79\\
D&60-69\\
F&59 or below
\end{tabular}
\end{tabular}
\end{center}

The instructor reserves the right to deviate from the above percentages in special cases, including borderline cases (generally this could be +/- 3\% points) may be given a + or − within the above intervals or increasing the letter grade.

\section*{Diversity Statement}
The University of Nevada, Reno is committed to providing a safe learning and work environment for all. If you believe you have experienced discrimination, sexual harassment, sexual assault, domestic/dating violence, or stalking, whether on or off campus, or need information related to immigration concerns, please contact the University’s Equal Opportunity \& Title IX Office at (775) 784-1547. Resources and interim measures are available to assist you. For more information, please visit \url{http:www.unr.edu/equal-opportunity-title-ix}.

\section*{Disability Statement}
The Department of Mathematics and Statistics supports providing equal access for students with disabilities. Any student with a disability needing academic adjustments or accommodations is requested to speak with me or the Disability Resource Center (PSAC 230, \url{http:www.unr.edu/drc}) as soon as possible to arrange for appropriate accommodations.

\section*{Academic Conduct}
No laptops, cell phones, mp3 players, or other electronics are to be used for personal reasons in class. If you are being disruptive during class you will be asked to leave. Disruptions in this context include inadequate participation. You must come to class on time and stay until the end of lecture. Tardy students will not be admitted to class. Please visit \url{http:www.unr.edu/student-conduct} for our official student code of conduct.

\section*{Academic Success Services}
A common habit among successful students is to seek help outside of the classroom. Your student fees cover use of the Math Center (784-4433 or \url{http:www.unr.edu/mathcenter}), Tutoring Center (784-6801 or \url{http:www.unr.edu/tutoring-center}), and University Writing Center (784-6030 or \url{http:www.unr.edu/writing-center}). These centers support your classroom learning; it is your responsibility to take advantage of their services.

\section*{University Recording Policy}
Surreptitious or covert videotaping of class or unauthorized audio recording of class is prohibited by law and by Board of Regents policy. This class may be videotaped or audio recorded only with the written permission of the instructor. In order to accommodate students with disabilities, some students may have been given permission to record class lectures and discussions. Therefore, students should understand that their comments during class may be recorded.

\section*{Academic Dishonesty}
Cheating, plagiarism, or otherwise obtaining grades under false pretenses constitutes academic dishonesty according to the code of this university. Academic dishonesty will not be tolerated and penalties can include canceling a student’s enrollment without a grade or giving an F for the assignment or for the entire course. For more details, see the University of Nevada, Reno general catalog.


%% in my own book schedule, I'd introduce sampling posterior much earlier

\section*{Tentative course schedule}
\begin{center}
  \begin{tabular}{|c|c|c|c|}
    \hline
    Week & Topic1 & Topic2 & Notes \\
    \hline
    1 & Syllabus/expectations/tools & Basic data structures &  \\
    \hline
    2 & \textbf{Labor Day (no class)} & Indexing/Iteration &  \\
    \hline
    3 & Text manipulation & Lab & \\
    \hline
    4 & Data frames and apply & Lab &  \\
    \hline
    5 & Review session & \textbf{Midterm 1} & \\
    \hline
    6 & Data Viz & Lab  & \\
    \hline
    7 & Functions & Lab & \\
    \hline
    8 & Simulation (Monte Carlo) & Lab & \\
    \hline
    9 & Importing data & Lab & \\
    \hline
    10 & Review session & \textbf{Midterm 2} & \\
    \hline
    11 & Split-apply-combine/parallelization & Lab & \\
    \hline
    12 & \textbf{Veteran's Day (no class)} & Fitting models/prediction & \\
    \hline
    13 & Monte Carlo integration/inference & lab & \\
    \hline
    14 & Bootstrap/jackknife & Lab & \\
    \hline
    15 & Optimization/Maximum likelihood estimation (MLE) & Lab & \\
    \hline
    16 & Markov Chain Monte Carlo (MCMC) & Lab/\textbf{presentations} & \\
    \hline
    17 & & \textbf{Final exam} & \\
    \hline
\end{tabular}
\end{center}

\end{document}