---
title: "Lab 1: R Basics"
author: "Intro to Statistical Computing, STAT 445-645"
date: "Wednesday, August 28, 2019"
---

Name:
NSHE ID:
Collaborated with:

This lab is to be completed both within and outside class. You're encouraged collaborate with your classmates, but you must identify their names above, and you must submit **your own** lab as an knitted HTML file on Canvas, by Sunday, September 15, 2019 by 11:59pm.

```{r}
## For reproducibility --- don't change this!
set.seed(09102018)
```

**This week's agenda**: manipulating data objects; using built-in functions, doing numerical calculations, and basic plots; reinforcing core probabilistic ideas.

The binomial distribution
===

The binomial distribution $\mathrm{Bin}(m,p)$ is defined by the number of successes in $m$ independent trials, each have probability $p$ of success. Think of flipping a coin $m$ times, where the coin is weighted to have probability $p$ of landing on heads.

The R function `rbinom()` generates random variables with a binomial distribution. E.g., 

```{r,eval=FALSE}
rbinom(n=20, size=10, prob=0.5)
```

produces 20 observations from $\mathrm{Bin}(10,0.5)$.

Some simple manipulations
===

- **1a.** Generate 200 random values from the $\mathrm{Bin}(10,0.5)$ distribution, and store them in a vector called `bin_draws_0.5`. Extract and display the first 10 elements. Extract and display all but the first 175 elements. 

- **1b.** Add the first element of `bin_draws_0.5` to the seventh. Compare the second element to the fifth, which is larger? A bit more tricky: print the indices of the elements of `bin.draws.0.5` that are equal to 5. How many such elements are there? **Challenge**: theoretically, how many such elements would you expect there to be?

- **1c.** Find the mean and standard deviation of `bin_draws_0.5`. Is the mean close what you'd expect? The standard deviation?

- **1d.** Call `summary()` on `bin_draws_0.5` and describe the result.

- **1e.** Find the data type of the elements in `bin_draws_0.5` using `typeof()`. Then convert `bin_draws_0.5` to a vector of characters, storing the result as `bin_draws_0.5_char`, and use `typeof()` again to verify that you've done the conversion correctly. Call `summary()` on `bin_draws_0.5_char`. Is the result formatted differently from what you saw above? Why?

Some simple plots
===

- **2a.** The function `plot()` is a generic function in R for the visual display of data. The function `hist()` specifically produces a histogram display. Use `hist()` to produce a histogram of your random draws from the binomial distribution, stored in `bin_draws_0.5`. 

- **2b.** Call `tabulate()` on `bin_draws_0.5`. What is being shown? Does it roughly match the histogram you produced in the last question?

- **2c.** Call `plot()` on `bin_draws_0.5` to display your random values from the binomial distribution. Can you guess what the `plot()` function is doing here?

- **2d.** Call `plot()` with two arguments, the first being `1:200`, and the second being `bin_draws_0.5`. This creates a scatterplot of `bin_draws_0.5` (on the y-axis) versus the indices 1 through 200 (on the x-axis). Does this match your plot from the last question?

More binomials, more plots
===

- **3a.** Generate binomial random variables again, but now produce 7 different vectors of length 200, with probability of success equal to: 0.2, 0.3, 0.4, 0.6, 0.7, and 0.8, storing the results in vectors called `bin_draws_0.2`, `bin_draws_0.3`, `bin_draws_0.4.`, `bin_draws_0.6`, `bin_draws_0.7` and  `bin_draws_0.8`. For each, compute the mean and standard deviation.

- **3b.** We'd like to compare the properties of our vectors. Create a vector of length 7, whose entries are the means of the 7 vectors we've created, in order according to the success probabilities of their underlying binomial distributions (0.2 through 0.8).

- **3c.** Using the vectors from the last part, create the following scatterplots. Explain in words, for each, what's going on.    

    * The 7 means versus the 7 probabilities used to generate the draws.
    * The standard deviations versus the probabilities. 
    * The standard deviations versus the means.
    
    **Challenge**: for each plot, add a curve that corresponds to the relationships you'd expect to see in the theoretical population (i.e., with an infinite amount of draws, rather than just 200 draws).

Working with matrices
===

- **4a.** Create a matrix of dimension 200 x 7, called `bin_matrix`, whose columns contain the 7 vectors we've created, in order of the success probabilities of their underlying binomial distributions (0.2 through 0.8). 

- **4b.** Print the first three rows of `bin_matrix`. Print the element in the 77th row and 6th column. Compute the largest element in first column. Compute the largest element in all but the first column.

- **4c.** Calculate the means of each column by calling `colMeans()` on `bin_matrix`. 

- **4d.** Compare the means you computed in the last question to those you computed in Q3b. Do they appear equal, when you print the results out to the console? Try using `identical()` to check that they are exactly equal. Does it fail? Why? Try using `all.equal()`. What does it report?

- **4e.** Take the transpose of `bin_matrix` and then take row means. Are these the same as what you just computed? Should they be?

Warm up is over, let's go big
===

- **5a.** R's capacity for data storage and computation is very large compared to what was available 10 years ago. Generate 2 million numbers from $\mathrm{Bin}(1 \times 10^6, 0.5)$ distribution and store them in a vector called `big_bin_draws`. Calculate the mean and standard deviation of this vector.

- **5b.** Create a new vector, called `big_bin_draws_standardized`, which is given by taking `big_bin_draws`, subtracting off its mean, and then dividing by its standard deviation. Calculate the mean and standard deviation of `big_bin_draws_standardized`. (These should be 0 and 1, respectively, or very, very close to it; if not, you've made a mistake somewhere).

- **5c.** Plot a histogram of `big_bin_draws_standardized`. To increase the number of histogram bars, set the `breaks` argument in the `hist()` function (e.g., set `breaks=100`). What does the shape of this histogram appear to be? Is this surprising? What could explain this phenomenon? (Hint: rhymes with "Mental Gimmick Serum" ...)

- **5d.** Calculate the proportion of times that an element of `big_bin_draws_standardized` exceeds 1.644854. Is this close to 0.05? 

- **5e.** Either by simulation, or via a built-in R function, compute the probability that a standard normal random variable exceeds 1.644854. Is this close to 0.05? (Hint: for either approach, it would be helpful to look at the help file for the `rnorm()` function.)

Now let's go really big
===

- **6a.** Let's push R's computational engine a little harder. Design an expression to generate 100 million numbers from $\mathrm{Bin}(10 \times 10^6, 50 \times 10^{-8})$, to be saved in a vector called `huge_bin_draws`, but do not evaluate this command yet. Then ask your lab partner to name three of Justin Bieber's songs and simultaneously evaluate your R command that defines `huge_bin_draws`. Which finished first, R or your partner?  (Note: your partner cannot really win this game. Even if he/she finishes first, he/she still loses.)

- **6b.** Calculate the mean and standard deviation of `huge_bin_draws`. Are they close to what you'd expect? (They should be very close.) Did it take longer to compute these, or to generate `huge_bin_draws` in the first place?

- **6c.** Calculate the median of `huge_bin_draws`. Did this median calculation take longer than the calculating the mean? Is this surprising? 

- **6d.** Calculate *e* to the power of the median of the logs of `huge_bin_draws`, in one line of code. Did this take longer than the median calculation applied to `huge_bin_draws` directly? Is this surprising?

- **6e.** Plot a histogram of of `huge_bin_draws`, again with a large setting of the `breaks` argument (e.g., `breaks=100`). Describe what you see; is this different from before, when we had 2 million draws? **Challenge**: is this surprising? What distribution is this?

Going big with lists
===

- **7a.** Convert `big_bin_draws` into a list using `as.list()` and save the result as `big_bin_draws_list`. Check that you indeed have a list by calling `class()` on the result. Check also that your list has the right length, and that its 1049th element is equal to that of `big_bin_draws`.

- **7b.** Run the code below, to standardize the binomial draws in the list `big_bin_draws_list`. Note that `lapply()` applies the function supplied in the second argument to every element of the list supplied in the first argument, and then returns a list of the function outputs. Did this `lapply()` command take longer to evaluate than the code you wrote in Q5b? (It should have; otherwise your previous code could have been improved, so go back and improve it.) Why do you think this is the case?

```{r, eval=FALSE}
big_bin_draws_mean = mean(big_bin_draws)
big_bin_draws_sd = sd(big_bin_draws)
standardize = function(x) {
  return((x - big_bin_draws_mean) / big_bin_draws_sd)
}
big_bin_draws_list_standardized_slow = lapply(big_bin_draws_list, standardize)
```

- **7c.** Run the code below, which again standardizes the binomial draws in the list `big_bin_draws_list`, using `lapply()`. Why is it so much slower than the code in the last question? Think about what is happening each time the function is called.

```{r, eval=FALSE}
big_bin_draws_mean = mean(big_bin_draws)
big_bin_draws_sd = sd(big_bin_draws)
standardize_slow = function(x) {
  return((x - mean(big_bin_draws)) / sd(big_bin_draws))
}
big_bin_draws_list_standardized_slow = lapply(big_bin_draws_list, standardize)
```

- **7d.** Lastly, let's look at memory useage. The command `object.size(x)` returns the number of bytes used to store the object `x` in your current R session. Find the number of bytes used to store `big_bin_draws` and `big_bin_draws_list`. How many megabytes (MB) is this, for each object? Which object requires more memory, and why do you think this is the case? Remind yourself: why are lists special compared to vectors, and is this property important for the current purpose (storing the binomial draws)?
