---
title: "Sample Lab Report"
subtitle: "Subtitle if Needed"
author: "Name Of Student"
date: "9/17/2019"
output: 
  html_document:
    theme: cosmo
    highlight: tango
---

See [here](https://yihui.name/knitr/options/) for a list of available knitr chunk options or a description of the ones below.

<!-- This should be at the top of every Rmarkdown document -->

```{r, setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  fig.align = 'center',
  comment = NA,
  message = FALSE,
  warning = FALSE,
  error = FALSE,
  results = 'hold')

# I also like to load any necessary libraries in the setup chunk

# loads a set of packages: ggplot, tibble, dplyr, etc. I use this package
# mainly for plotting, working with data, and pipe operations (see later)
library(tidyverse)

# This package allows for creating nice looking tables from data frames and
# matrices
library(kableExtra)
```

<!-- the <br /> tag breaks the line without adding space between lines -->
**Name:** Look you get to write your name twice! <br />
**NSHE ID:** 000000 <br />
**Collaborated with:** N/A

Some short blurb about the lab and its contents. This can be one or a few sentences that you write or grabbed from the lab.

<!-- Two hashes to designate a section -->
## Section Header

Here is some background and setup information that pertains to the following questions.

<!-- Three hashes to designate a subsection -->
### Subsection header pertaining to tasks

<!--
You don't need to change the question text unless you want to cut out
unnecesary text. Ie cutting it down to just the task without hints, notes, etc.
-->
- **1a.** This is question **1a** telling you what to do. It may be a couple of sentences or it may be just one.

```{r, q1a}
n <- 10000
d <- 6

x <- rnorm(n * d)
normal_mat <- matrix(x, nrow = n, ncol = d)
head(normal_mat, 4)
```

- <span style="color:red; font-weight:bold">Answer 1a</span> This is my answer to the above question, or a comment about the output of the code. Your eyes are drawn right to this text because of the way it stands out against the question.

## Working with Data Frames

*Data frames are a special type of list where all vectors in the list are the same length*

### Showing Data Frames in Rmarkdown

- **2a.** Load the `iris` data set and print the first 5 rows using `head()`.

- <span style="color:red; font-weight:bold">Answer 2a</span> This is my explanation or statement about the code below.

```{r, q2a, echo=FALSE}
head(iris, 5)
```


- **2b.** Print the first 5 rows using `kableExtra`

- <span style="color:red; font-weight:bold">Answer 2b</span> This code also uses the *pipe* operator that is automatically loaded when `tidyverse` is loaded. Alternatively you can load it from the `magritr` package. Details about the `kableExtra` package can be found [here](https://haozhu233.github.io/kableExtra/awesome_table_in_pdf.pdf).

```{r, q2b}
n_rows <- 5
iris %>%
  head(n_rows) %>%
  kable(format = "html", digits = 2, align = 'c', booktabs = TRUE,
        caption = paste("These are the first", n_rows, "rows \
                        of the iris data set")) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))

# I try to keep the maximum number of characters per line at 80. Usually longer
# lines of code are harder to read or don't display as well on smaller devices.
```


## Fancy Markdown and Latex

Rmarkdown (and knitr) have the ability to render some fancy items, most notably being $\LaTeX$ code used for math. You can read more about LaTeX (and MathJax) for math type setting at the following sources

- [TeX Commands available in MathJax](http://www.onemathematicalcat.org/MathJaxDocumentation/TeXSyntax.htm)
- [MathJax basic tutorial and quick reference](https://math.meta.stackexchange.com/questions/5020/mathjax-basic-tutorial-and-quick-reference)

### Writing equations

Inline math statements can be put between single dollar signs \$ .. \$ so that it flows with the text. Commands such as `$\hat{\beta}_0$` will render as $\hat{\beta}_0$ without breaking the flow. Display math can be put in double dollar signs `$$ .. $$`. E.g.

$$
f(x | \mu, \sigma) = \frac{1}{\sqrt{2\pi\sigma^2}}\exp\left\lbrace\frac{-(x - \mu)^2}{2\sigma^2}\right\rbrace
$$

Multiline equations can be written using pure $LaTeX$ that gets rendered by knitr. The following code **IS NOT** between double dollar signs.

\begin{align}
f(x) &= ax^2 + bx + c \\
&= a\left(x^2 + \frac{b}{a}x\right) + c \\
&= a\left(x^2 + \frac{b}{a}x + \left(\frac{b}{2a}\right)^2\right) + c - a\left(\frac{b}{2a}\right)^2 \\
&= a\left(x + \frac{b}{2a}\right)^2 + \frac{4ac}{4a} - \frac{b^2}{4a} \\
&= a\left(x + \frac{b}{2a}\right)^2 + \frac{4ac - b^2}{4a}
\end{align}

## Setting Code Chunk Options

### Displaying Results

Consider the following (unevaluated) code:

```{r, eval=FALSE}
x1 <- rnbinom(100, 5, 0.1)
(x1.mean <- mean(x1))
(x1.sd <- sd(x1))
x2 <- rnbinom(100, 5, 0.2)
(x2.mean <- mean(x2))
(x2.sd <- sd(x2))
x3 <- rnbinom(100, 5, 0.3)
(x3.mean <- mean(x3))
(x3.sd <- sd(x3))
x4 <- rnbinom(100, 5, 0.4)
(x4.mean <- mean(x4))
(x4.sd <- sd(x4))
x5 <- rnbinom(100, 5, 0.5)
(x5.mean <- mean(x5))
(x5.sd <- sd(x5))
x6 <- rnbinom(100, 5, 0.6)
(x6.mean <- mean(x6))
(x6.sd <- sd(x6))
x7 <- rnbinom(100, 5, 0.7)
(x7.mean <- mean(x7))
(x7.sd <- sd(x7))
x8 <- rnbinom(100, 5, 0.8)
(x8.mean <- mean(x8))
(x8.sd <- sd(x8))
x9 <- rnbinom(100, 5, 0.9)
(x9.mean <- mean(x9))
(x9.sd <- sd(x9))
```

- There is a lot of repetitive code
- There is a lot of repetitive output

**results='markup'** *(default behavior, ugly)*

This is hard to look at and hard to read. It's also hard to find any pattern in the results.

```{r, results='markup'}
x1 <- rnbinom(100, 5, 0.1)
(x1.mean <- mean(x1))
(x1.sd <- sd(x1))
x2 <- rnbinom(100, 5, 0.2)
(x2.mean <- mean(x2))
(x2.sd <- sd(x2))
x3 <- rnbinom(100, 5, 0.3)
(x3.mean <- mean(x3))
(x3.sd <- sd(x3))
x4 <- rnbinom(100, 5, 0.4)
(x4.mean <- mean(x4))
(x4.sd <- sd(x4))
x5 <- rnbinom(100, 5, 0.5)
(x5.mean <- mean(x5))
(x5.sd <- sd(x5))
x6 <- rnbinom(100, 5, 0.6)
(x6.mean <- mean(x6))
(x6.sd <- sd(x6))
x7 <- rnbinom(100, 5, 0.7)
(x7.mean <- mean(x7))
(x7.sd <- sd(x7))
x8 <- rnbinom(100, 5, 0.8)
(x8.mean <- mean(x8))
(x8.sd <- sd(x8))
x9 <- rnbinom(100, 5, 0.9)
(x9.mean <- mean(x9))
(x9.sd <- sd(x9))
```

**results='hold'** *(good)*

A lot of repetitive code that's only calculating simple things. Probably not worth showing the output of such a task.

```{r, results='hold'}
x1 <- rnbinom(100, 5, 0.1)
(x1.mean <- mean(x1))
(x1.sd <- sd(x1))
x2 <- rnbinom(100, 5, 0.2)
(x2.mean <- mean(x2))
(x2.sd <- sd(x2))
x3 <- rnbinom(100, 5, 0.3)
(x3.mean <- mean(x3))
(x3.sd <- sd(x3))
x4 <- rnbinom(100, 5, 0.4)
(x4.mean <- mean(x4))
(x4.sd <- sd(x4))
x5 <- rnbinom(100, 5, 0.5)
(x5.mean <- mean(x5))
(x5.sd <- sd(x5))
x6 <- rnbinom(100, 5, 0.6)
(x6.mean <- mean(x6))
(x6.sd <- sd(x6))
x7 <- rnbinom(100, 5, 0.7)
(x7.mean <- mean(x7))
(x7.sd <- sd(x7))
x8 <- rnbinom(100, 5, 0.8)
(x8.mean <- mean(x8))
(x8.sd <- sd(x8))
x9 <- rnbinom(100, 5, 0.9)
(x9.mean <- mean(x9))
(x9.sd <- sd(x9))
```

**using a matrix or data frame** *(better)*

Code is more concise. We can now see a pattern! As the probability goes up, the mean and standard deviation go down.

```{r}
x1 <- rnbinom(100, 5, 0.1)
x2 <- rnbinom(100, 5, 0.2)
x3 <- rnbinom(100, 5, 0.3)
x4 <- rnbinom(100, 5, 0.4)
x5 <- rnbinom(100, 5, 0.5)
x6 <- rnbinom(100, 5, 0.6)
x8 <- rnbinom(100, 5, 0.8)
x7 <- rnbinom(100, 5, 0.7)
x9 <- rnbinom(100, 5, 0.9)

X <- cbind(x1, x2, x3, x4, x5, x6, x7, x8, x9)
X.summary <- data.frame(Prob = seq(0.1, 0.9, 0.1),
                        Mean = colMeans(X),
                        `Std.Dev` = apply(X, 2, sd))

X.summary
```

**parametric code + matrix / data frame + kable** *(best)*

The best code is concise, scalable, and easy to understand. 

Also consider: what if I want to go and add another statistic (say the median)? With the first method, I would have to type out `xN.median <- median(xN)` 9 times! Or what if I want to be more precise with the probability step sizes (eg every 0.05 steps). Now that's over 40 lines of extra code that I need to write!

The amount of code you have to write increases very quickly, whereas with the code below, I would only need to change `seq(0.1, 0.9, 0.1)` to `seq(0.1, 0.9, 0.05)` and add one more line to the `data.frame`.

```{r}
n <- 100
size <- 5
probs <- seq(0.1, 0.9, 0.1)

X <- sapply(probs, function(prob){
  rnbinom(n = n, size = size, prob = prob)
})

X.summary <- data.frame(Probability = seq(0.1, 0.9, 0.1),
                        Mean = colMeans(X),
                        `Std.Dev` = apply(X, 2, sd))
```

```{r}
# This chunk can/should be hidden because it is only used for display output
X.summary %>%
  kable(digits = 3, align = 'r',
        caption = "Mean and Std. Dev. of different Negative Binomial distributions") %>%
  kable_styling(full_width = TRUE, bootstrap_options = "striped")
```

