---
title: "Data Structures in R"
author: "AG Schissler"
date: "8/29/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## I. Opening 

### Start-of-class work

1. Log in the your workstation.
2. Open the RStudio application on your machine.

### Today learning outcomes

Students will be able to

- use R vectors.
- use R matrices.
- use R factors.
- use R lists.
- use data.frames.

### Some things I'd like to highlight in R 

#### Data types

At base level, all data can represented in binary format, by **bits** (i.e., TRUE/FALSE, YES/NO, 1/0). Basic data types:

- **Booleans**: Direct binary values: `TRUE` or `FALSE` in R
- **Integers**: whole numbers (positive, negative or zero), represented by a fixed-length block of bits
- **Characters**: fixed-length blocks of bits, with special coding; **strings**: sequences of characters
- **Floating point numbers**: an integer times a positive integer to the power of an integer, as in $3 \times 10^6$ or $1 \times 3^{-1}$
- **Missing or ill-defined values**: `NA`, `NaN`, etc.

#### Operators
- **Unary**: take just one argument. E.g., `-` for arithmetic negation, `!` for Boolean negation
- **Binary**: take two arguments. E.g., `+`, `-`, `*`, and `/` (though this is only a partial operator). Also, `%%` (for mod), and `^` (again partial)

```{r}
-7
7 + 5
7 - 5
```

```{r}
7 * 5
7 ^ 5
7 / 5
7 %% 5
```

#### R workspace

What variables have you defined?

```{r}
ls()
```

Getting rid of variables:

```{r}
x_vec <- y_vec <- 1:5
ls()
rm("x_vec")
ls()
rm(list=ls()) # Be warned! This erases everything
ls()
```

## II. First data structure: Vectors

- A **data structure** is a grouping of related data values into an object
- A **vector** is a sequence of values, all of the same type

### I do

```{r}
x = c(7, 8, 10, 45)
x
is.vector(x)
```

- The `c()` function returns a vector containing all its arguments in  specified order
- `1:5` is shorthand for `c(1,2,3,4,5)`, and so on
- `x[1]` would be the first element, `x[4]` the fourth element, and `x[-4]` is a vector containing *all but* the fourth element

---

`vector(length=n)` returns an empty vector of length *n*; helpful for filling things up later

```{r}
weekly.hours = vector(length=5)
weekly.hours
weekly.hours[5] = 8
weekly.hours
```
---
Arithmetic operator apply to vectors in a "componentwise" fashion

```{r}
y = c(-7, -8, -10, -45)
x + y
x * y
```
---
**Recycling** repeat elements in shorter vector when combined with a longer one

```{r}
x + c(-7,-8)
x^c(1,0,-1,0.5)
```

Single numbers are vectors of length 1 for purposes of recycling:

```{r}
2 * x
```

### We do

#### Apply some base (built-in) R functions to `x` and `y`.

Many functions can take vectors as arguments:

- `mean()`, `median()`, `sd()`, `var()`, `max()`, `min()`, 
`length()`, and `sum()` return single numbers
- `sort()` returns a new vector
- `hist()` takes a vector of numbers and produces a histogram, 
a highly structured object, with the side effect of making a plot
- `ecdf()` similarly produces a cumulative-density-function object
- `summary()` gives a five-number summary of numerical vectors
- `any()` and `all()` are useful on Boolean vectors

```{r}
## Our code here

```

### You do

#### Using R to be fast like Gauss

1. Define a variable `x_vec` to contain the integers 1 through 100. 
2. Check that it has length 100.
3. Report the data type being stored in `x_vec`.
4. Add up the numbers in `x_vec`, by calling a built-in R function. 
5. How many arithmetic operations did this take?  
**Challenge**: Show how Gauss would have done this same calculation as a 7 year old, using just 3 arithmetic operations.

## III. Using Matrices in R

### We do

1. Convert `x_vec` into a matrix with 20 rows and 5 columns, and store this as `x_mat`. Here `x_mat` should be filled out in the default order (column major order). 
3. Check the dimensions of `x_mat`, and the data type as well. 
4. Compute the sums of each of the 5 columns of `x_mat`, by calling a built-in R function. 

### You do

1. Extract and display rows 1, 5, and 17 of `x_mat`, with a single line of code. Answer the following questions, each with a single line of code: how many elements in row 2 of `x_mat` are larger than 40? How many elements in column 3 are in between 45 and 50? How many elements in column 5 are odd? Hint: take advantage of the `sum()` function applied to Boolean.__

2. Using Boolean indexing, modify `x_vec` so that every even number in this vector is incremented by 10, and every odd number is left alone. This should require just a single line of code. Print out the result to the console.__

**Challenge**: show that `ifelse()` can be used to do the same thing, again using just a single line of code.

## IV. Third data structure: lists

A **list** is sequence of values, but not necessarily all of the same type

### Key-value pairs

- Lists give us a natural way to store and look up data by _name_, rather than by _position_
- A really useful programming concept with many names: **key-value pairs**, i.e., **dictionaries**, or **associative arrays**.
- If all our distributions have components named `family`, we can look that up by name, without caring where it is (in what position it lies) in the list.

### I do

Consider the list `x.list` created below. 

```{r}
set.seed(08292018)
x_list = list(rnorm(6), letters, sample(c(TRUE,FALSE),size=4,replace=TRUE))
```
### We do

Complete the following tasks, each with a single line of code: 
1. Extract all but the second element of `x_list`---seeking here a list as the final answer. 
2. Extract the first and third elements of `x_list`, then extract the second element of the resulting list---seeking here a vector as the final answer. 

```{r}
## Our code here.
```

### You do

Extract the second element of `x_list` as a vector, and then extract the first 10 elements of this vector---seeking here a vector as the final answer. Note: pay close attention to what is asked and use either single brackets `[ ]` or double brackets ``[[ ]]`` as appropriate.

## V. Fifth data structure: `data.frame` or really just a list?

### I do

### We do

### You do

## VI. Closing: Mind map

Mind mapping is a powerful learning strategy that uses the associative powers of your mind, coupled with imagery and verbal labeling.

See the [wiki](https://en.wikipedia.org/wiki/Mind_map) page.

### I do
I'll demonstrate the construction of a "mind map" (aka "concept map") from the first lesson.

### You do
Please create a on today's lesson.

### We do
Share insights from the process with the class.
